---
title: "Spotify Music Analysis"
author: "Shammah"
date: "2025-02-18"
output: html_document
---

- using dataset from [leonardopena/top-spotify-songs-from-20102019-by-year](https://www.kaggle.com/datasets/leonardopena/top-spotify-songs-from-20102019-by-year)

- Description of headings [playlistmachinery](http://organizeyourmusic.playlistmachinery.com/#)


1. **Genre** - the genre of the track
2. **Year** - the release year of the recording. Note that due to vagaries of releases, re-releases, re-issues and general madness, sometimes the release years are not what you'd expect.
3. **Added** - the earliest date you added the track to your collection.
4. **Beats** Per Minute (BPM) - The tempo of the song.
5. **Energy** - The energy of a song - the higher the value, the more energtic. song
6. **Danceability** - The higher the value, the easier it is to dance to this song.
7. **Loudness (dB)** - The higher the value, the louder the song.
8. **Liveness** - The higher the value, the more likely the song is a live recording.
9. **Valence** - The higher the value, the more positive mood for the song.
10. **Length** - The duration of the song.
11. **Acousticness** - The higher the value the more acoustic the song is.
12. **Speechiness** - The higher the value the more spoken word the song contains.
14. **Popularity** - The higher the value the more popular the song is.
15. **Duration** - The length of the song. 

```{r}
library(data.table)

url <- "./data/top10s.csv"
data <- fread(input = url, header = TRUE)
head(data)
```
```{r}
str(data)
```
```{r}
summary(data)
```
```{r}
colnames(data)
```
```{r}
song_genre_df <- data.frame(data[ ,c("title", "top genre")])
head(song_genre_df)
```
```{r}
genre_vector <- c(song_genre_df[ , "top.genre"])
#genre_vector
#genre_factor <- as.factor(genre_vector)
#str(genre_factor)
genre_freq_table <- data.frame(table(data$`top genre`))
head(genre_freq_table)
```
```{r}
sorted_by_genre_freq <- genre_freq_table[order(genre_freq_table$Freq, decreasing = TRUE), ]
head(sorted_by_genre_freq)
```

