---
title: "Spotify Music Analysis"
author: "Shammah"
date: "2025-02-18"
output: html_document
---

## Introduction
The main goal of this analysis is to figure out what factors contribute to the
popularity of a song from multiple variables. Songs in this data set were popular
enough to be featured on the billboard top 100 songs list at different times. 

We are using dataset from [leonardopena/top-spotify-songs-from-20102019-by-year](https://www.kaggle.com/datasets/leonardopena/top-spotify-songs-from-20102019-by-year)

### Description of Columns [playlistmachinery](http://organizeyourmusic.playlistmachinery.com/#)


1. **Genre** - the genre of the track
2. **Year** - the release year of the recording. Note that due to vagaries of releases, re-releases, re-issues and general madness, sometimes the release years are not what you'd expect.
3. **Added** - the earliest date you added the track to your collection.
4. **Beats** Per Minute (BPM) - The tempo of the song.
5. **Energy** - The energy of a song - the higher the value, the more energtic. song
6. **Danceability** - The higher the value, the easier it is to dance to this song.
7. **Loudness (dB)** - The higher the value, the louder the song.
8. **Liveness** - The higher the value, the more likely the song is a live recording.
9. **Valence** - The higher the value, the more positive mood for the song.
10. **Length** - The duration of the song.
11. **Acousticness** - The higher the value the more acoustic the song is.
12. **Speechiness** - The higher the value the more spoken word the song contains.
14. **Popularity** - The higher the value the more popular the song is.
15. **Duration** - The length of the song. 

### Exploring the Data
```{r}
library(data.table)

url <- "./data/top10s.csv"
data <- as.data.frame(fread(input = url, header = TRUE))
colnames(data)[colnames(data) == "top genre"] <- "genre"
head(data)
tail(data)
```
Song rankings in the data set are based on the Popularity (pop column). We will be
evaluating all other variables in relation to the Popularity.
```{r}
str(data)
```
The dataset contains 15 variables of which 9 of them are more relevant for this analysis.
The 9 relevant variables are based on song characteristics like bpm, genre and loudness.
```{r}
summary(data)
```
This dataset does not have missing values.
```{r}
colnames(data)
```

#### Checking for Posible Groups
```{r}
genre_vector <- c(data$genre)
genre_factor <- as.factor(genre_vector)
str(genre_factor)
```

Genre is typically a good variable to classify music on. In this dataset however,
we have 50 genres which can be a bit challenging to work with. We will create a 
frequency table for the genre to figure out if there are enough data points to group
our songs by.

```{r}
genre_freq_table <- data.frame(table(data$genre))
head(genre_freq_table)
```


```{r}
sorted_by_genre_freq <- genre_freq_table[order(genre_freq_table$Freq, decreasing = TRUE), ]
head(sorted_by_genre_freq, n=10)
```
Sorting the genre frequency table shows a clear distinction among different genres.
Based on the frequency of a genre, we can determine which genre is more likely to
rank high enough to be included in the bill board top 100.

### Investigating Correlation Genre/Populatiry
Now that we know some genres have higher frequency on the bill board top 100 with 
dance pop songs showing up **327** times, we will investigate if the genre has any impact on popularity.

We will use the **Shapiro Wilk** test to determine if the popularity if normally distributed
If we have normal distribution, we can use the **ANOVA test** (Analysis of Variances)
otherwise we will use the **Kruskal-Wallis test**

**ANOVA test** is used on normal distribution to check for significant differences 
between the means of two or more independent variables, in this case *genre* and *pop*.

**Kruskal-Wallis test** is a non-parametric alternative to ANOVA test used to compare the
medias of those same variables.

A *p-value* < 0.05 on the Shapiro-Wilk test means the data is not in a normal distribution
hence the Krustal-Wallis test is recommended.

The *p-value* from Kruskal-Wallis* is < 0.05 indicating at least one genre has
significantly popularity which the specific genre comparisons can be done using 
**Dunn's post-hoc test** which we can look at as an addition to the current scope of this analysis.

We will add a *Spearman test* to get a correlation value. *P-value* < 0.05 shows significant correlation
*rho*>0 shows a positive correlation between the genre and popularity.
```{r}
data$genre_numeric <- as.integer(factor(data$genre))
head(data)
shapiro.test(data$pop)
kruskal.test(pop ~ genre, data = data)
cor.test(data$pop, data$genre_numeric, method = "spearman")
```

### Visualization
```{r}
suppressPackageStartupMessages(library(ggcorrplot))

m <- cor(data[,c(6:16) ])

ggcorrplot(m, lab = TRUE, lab_size = 2,ggtheme = ggplot2::theme_minimal(), 
           method = "square", colors = c("blue", "white", "red"))
```
```{r}
suppressPackageStartupMessages(library(dplyr))
library(ggplot2)

dance_pop_data <- data %>% 
  filter(genre == "dance pop")

head(dance_pop_data)

ggplot(dance_pop_data, aes(x = as.factor(year))) +
  geom_bar(fill = "skyblue") +
  labs(title = "Count of Dance Pop Genre by Year", x = "Year", y = "Count of Occurrences") +
  theme_minimal()
```
```{r}
genre_year_count <- data %>%
  group_by(year, genre) %>%
  summarise(count = n(), .groups = 'keep')

head(genre_year_count)
options(repr.plot.width = 18, repr.plot.height = 8)


last_data_points <- genre_year_count %>%
  group_by(genre) %>%
  slice_max(year, n = 1)

ggplot(genre_year_count, aes(x = year, y = count, color = genre, group = genre)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, aes(color = genre),linewidth = 0.5, linetype = "dashed") +
  geom_text(data = last_data_points,
            aes(label = genre), 
            hjust = -0.1,  
            vjust = 0.5,
            check_overlap = TRUE)+
  labs(title = "Count of Genres by Year", x = "Year", y = "Count of Occurrences") +
  theme_minimal() +
  theme(legend.position = "none",
        legend.justification = "center")

ggplot(genre_year_count, aes(x = year, y = count, color = genre, group = genre)) +
  geom_line() +
  labs(title = "Count of Genres by Year", x = "Year", y = "Count of Occurrences") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

### Conclusion
Factors that have an impact on the popularity or a positive correlation are danceability,
loudness and genre. This analysis mainly focused on the genre which is usually more within
an artist's control. 

Dance Pop was the most popular genre from 2010 through 2018 before the popularity goes down
towards 2019. The sharp decline in Dance Pop songs on the bill board 100 list represents a 
large change in interest in the music market and we would not recommend new artists to jump
into that genre.

Pop music is trending upwards and would be a good recommendation for an artist looking to 
jump on to a positive trend. A few of the genres with smaller frequencies on the bill boards list
are also trending upwards. It would be worth while to investigate which particular artists are 
trending and how they are reaching their audiences to create a shift in the music market.